Missing the Point\(\colon\) Non-Convergence in Iterative Imputation Procedures

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Add example of non-convergence

Missing data problems are ubiquitous and pervasive in the social and
behavioral sciences. If a dataset contains just one missing value for a
variable of interest, statistical inferences are undefined and will not
produce any result. Incomplete observations are therefore ignored by
default in many statistical packages (i.e., list-wise deletion is
employed). Unfortunately, this \emph{ad hoc} solution may yield wildly
invalid results \citep{buur18}. An alternative is to `impute' (i.e.,
fill in) every missing value in an incomplete dataset. With imputation
techniques, one or several completed datasets are created, on which
statistical inferences can be performed. The case with several completed
datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
requires an additional step to pool the results across imputations
\textbf{see figure 1??}. By applying Rubin's rules, the pooled result
will reflect the uncertainty in the data due to missingness. Under many
circumstances, this yields an unbiased and confidence valid estimate of
the inference on the true--but missing--data \citep{buur18}. A popular
method to obtain imputations is to use the `Multiple Imputation by
Chained Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an
iterative algorithmic procedure to draw imputations from the posterior
predictive distribution of the missing values. This introduces a
potential threat to the validity of the imputations: what if the
algorithm has not converged? Are the implications then to be trusted?
And can we rely on the inference obtained on the completed data? These
are all open questions, because the convergence properties of iterative
imputation algorithms have not been systematically studied
\citep{buur18}. Moreover, there is no scientific consensus on the
convergence properties of MI algorithms \citep{taka17}. Some default
MICE techniques (e.g., `predictive mean modeling') might not yield
converged states at all \citep{murr18}. Therefore, algorithmic
convergence should be monitored carefully.

@@

Main question: How can non-convergence be diagnosed? Sub-questions: Are
common MCMC non-convergence diagnostics appropriate for MICE? And if so,
which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{(These are part of the methods, not
RQ per se: What are the effects of continued iterations on estimates,
predictions and inferences? And finally, do the answers differ with
varying missingness proportions? That is, we vary the nr of iterarions
and the missingness proportion because we assume that the algorithm has
not reached convergence at \(t=1\), and performs worse with moderate to
high missingness (not so much with little or a LOT of missingness).)}

@@

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}, see Figure 1. Without mixing,
imputation chains do not intermingle nicely. Without stationarity, there
is trending within imputation chains. While the current practice is to
inspect traceplots for signs of non-mixing and non-stationarity,
diagnostic evaluation of these two challenges would be preferred.
Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used.

@@

The two challenges of convergence are stationarity and mixing point.
Mixing and stationarity can be evaluated visually. Common practice is to
evaluate stationarity and mixing in trace plots for scalar summaries of
interest. In mice these scalar summaries of interest are chain means and
chain variances. As Van Buren describes users can also specify a model
specific scalar summary (e.g., a regression coefficient). search a user
specified scalar summary however is not universal to all complete data
problems. Therefore as inspired by someone 2002 we will use introduce a
scalar summary that summarizes the multivariate state space of the
algorithm. In this study we use the first eigenvalue of the variance
covariance matrix of the completed data set per imputation. The methods
under evaluation are Rhat and AC. These will be computed for the three
scalar summaries of interest: chain means, chain variances, and the
first eigenvalue of the variance-covariance matrix across imputations.

The two methods will be evaluated with an estimand per scalar summary of
interest. It is assumed that when the estimates are unbiased (and
confidence valid), the algorithm is sufficiently converged. The R hat
and AC values with chain means as theta are evaluated against the bias
in univariate mean estimates (\textbf{just for Y??}). Similarly, the
performance measure for Rhat and AC applied to chain variances is the
bias in estimated standard deviations. To evaluate the performance of
Rhat and AC on the eigenvalues, we will use bias in R squared and in the
estimated regression coefficient, the coverage rate of the CI95\% of the
regression estimate, and the CI length. \textbf{remove the term
performance measure above and add that performance measures are bias in
all estimands, and cov and CIL of regression coeff.}

@@

In general there is more bias in the estimate in conditions with a
higher proportion of missingness and or a lower number of iterations. As
required the values of or hat R lower for conditions with a higher
number off iterations. End somewhat higher 4 conditions with a higher
percentage of missing data. However autocorrelation did not show a
decrease with increase number of iterations. After replacing the ACF
function with a manual calculation for autocorrelation the
autocorrelation values were indeed decreasing with a higher number of
iterations. Evaluation with the performance measures shows that are had
an autocorrelation are conservative: signs of non-convergence in
conditions where estimands are unbiased and or confidence valid. As
expected complete convergence (as indicated by autocorrelation equals 0;
R hat equals 1) is not observed.

@@

Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
1.01. And a t-test is performed for AC values == 0. Performance measures
are the same as above: unbiased, confidence valid estimates.

@@

Our head is equal to 1.2 is not stringent enough because the performance
measures R not unbiased and confidence fellas fell it 4 every condition
in which our hat is smaller than one point 2. At least, not when our hat
is smaller than one point two fur first occurs. The threshold R hat is
equal to one point one is somewhat conservative with respect to the
performance measures. It seems OK. Our hat is equal to 1.01 is too
stringent compared to the performance measures (\textbf{but may be
necessary in more complex complete data models???}).

@@

For empirical researchers: 1) Check trace plots for pathological non
convergence and adjust imputation model if necessary. 2) use our hat
wait 1.1 ash threshold and autocorrelation with \textbf{?} as threshold
. Keep iterating until these thresholds are reached. 3) Do not use the R
function ACF. Instead compute autocorrelations manually , see for
example GitHub link. 4) Track your own scalar summary of interest. This
is advanced but explained in Van Buren 2018. Compute R hat and
autocorrelation values for this scalar summary.

@@

If I don't include it in my study, also: Significance of AC values;
Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2, ..., 100\)), and 2) varying the
missingness proportions (\(p_{miss} = .05, .25, .50, .75, .95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the inital values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

@@

\(\widehat{R}\) (\textbf{of what?? i.e., chain means and chain variances
of each variable}) is computed by implementing Vehtari et al.'s
\citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
is AC aggregated across imputations? i.e., mean in latest sim}) is
computed with function \texttt{stats::acf()}.

To estimate the quantity of scientific interest, \(Q\), we perform
multiple linear regression on each completed dataset with the function
\texttt{stats::lm()}. We obtain an estimated regression coefficient per
imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
the function \texttt{mice::pool()} to estimate u bar (\textbf{check wat
dat nou precies is!!}) according to Rubin's \citeyearpar{rubin87} rules,
and subsequently implement finite population pooling conform
\citet{vink14}.

@@

Auto-correlations indicate no sign of trending within imputation chains.
In most simulation conditions AC is smaller than or about equal to zero.
Across simulation conditions, however, the autocorrelation curve does
not trend towards 0. Autocorrelation values plateau off at a value of
around .1 . This is a small positive autocorrelation, which would
indicate some trending within chains.

@@

The negative \(AC\)-values obtained in this study show no threat of
non-stationarity. However, initial dip in \(AC\)-values may have
implications for the default number of iterations in \texttt{mice}
(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
the most appropriate, since this lead to the worst convergence
(\textbf{nope, only for Rh, not AC}), as indicated by \(\widehat{R}\)
and \(AC\). Under the current specifications, \(T>20\) would be more
appropriate.

The observed dip in AC implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant.

>>>>

Missing data problems are ubiquitous and pervasive in the social and
behavioral sciences. If a dataset contains just one missing value for a
variable of interest, statistical inferences are undefined and will not
produce any result. Incomplete observations are therefore ignored by
default in many statistical packages (i.e., list-wise deletion is
employed). Unfortunately, this \emph{ad hoc} solution may yield wildly
invalid results \citep{buur18}. An alternative is to `impute' (i.e.,
fill in) every missing value in an incomplete dataset. With imputation
techniques, one or several completed datasets are created, on which
statistical inferences can be performed. The case with several completed
datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
requires an additional step to pool the results across imputations
\textbf{see figure 1??}.

\textbf{Move this after notation??} Figure 1 provides an overview of the
steps involved with MI---from incomplete data, to \(M\) completed
datasets, to \(M\) estimated quantities of interest (\(\hat{Q}\)s), to a
single pooled estimate \(\bar{Q}\). Missing data in \(y\) is `imputed'
(i.e., filled in) \(M\) times. The imputed data (\(y_{imp}\)) is
combined with the observed data (\(y_{obs}\)) to create \(M\) completed
datasets. On each completed dataset, the analysis of scientific interest
(the `substantive model', or `complete data model') is performed. The
quantity of scientific interest (e.g., a regression coefficient) is
denoted with \(Q\). Since \(Q\) is estimated on each completed dataset,
\(M\) separate \(\hat{Q}\)-values are obtained. These \(M\) values are
combined into a single pooled estimate \(\bar{Q}\). \textbf{ALSO: make
clear that iteration happens within the first step.}

\begin{figure}

{\centering \includegraphics[width=\linewidth]{./images/diagram}

}

\caption{Scheme of the main steps in multiple imputation.}\label{fig:unnamed-chunk-1}
\end{figure}

By applying Rubin's rules, the pooled result will reflect the
uncertainty in the data due to missingness. Under many circumstances,
this yields an unbiased and confidence valid estimate of the inference
on the true--but missing--data \citep{buur18}. A popular method to
obtain imputations is to use the `Multiple Imputation by Chained
Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an iterative
algorithmic procedure to draw imputations from the posterior predictive
distribution of the missing values. This introduces a potential threat
to the validity of the imputations: what if the algorithm has not
converged? Are the implications then to be trusted? And can we rely on
the inference obtained on the completed data? These are all open
questions, because the convergence properties of iterative imputation
algorithms have not been systematically studied \citep{buur18}.
Moreover, there is no scientific consensus on the convergence properties
of MI algorithms \citep{taka17}. Some default MICE techniques (e.g.,
`predictive mean modeling') might not yield converged states at all
\citep{murr18}. Therefore, algorithmic convergence should be monitored
carefully.

@@

Main question: How can non-convergence be diagnosed? Sub-questions: Are
common MCMC non-convergence diagnostics appropriate for MICE? And if so,
which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{(These are part of the methods, not
RQ per se: What are the effects of continued iterations on estimates,
predictions and inferences? And finally, do the answers differ with
varying missingness proportions? That is, we vary the nr of iterations
and the missingness proportion because we assume that the algorithm has
not reached convergence at \(t=1\), and performs worse with moderate to
high missingness (not so much with little or a LOT of missingness).)}

@@

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}, see Figure XYZ. Without mixing,
imputation chains do not intermingle nicely. Without stationarity, there
is trending within imputation chains. While the current practice is to
inspect traceplots for signs of non-mixing and non-stationarity,
diagnostic evaluation of these two challenges would be preferred.
Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used.

\begin{figure}

{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/unnamed-chunk-2-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/unnamed-chunk-2-2}

}

\caption{Pathological non-convergence versus typical convergence.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

The two challenges of convergence are stationarity and mixing point.
Mixing and stationarity can be evaluated visually. Common practice is to
evaluate stationarity and mixing in trace plots for scalar summaries of
interest. In mice these scalar summaries of interest are chain means and
chain variances. As Van Buren describes users can also specify a model
specific scalar summary (e.g., a regression coefficient). search a user
specified scalar summary however is not universal to all complete data
problems. Therefore as inspired by someone 2002 we will use introduce a
scalar summary that summarizes the multivariate state space of the
algorithm. In this study we use the first eigenvalue of the variance
covariance matrix of the completed data set per imputation. The methods
under evaluation are \(\widehat{R}\) and \(AC\). These will be computed
for the three scalar summaries of interest: chain means, chain
variances, and the first eigenvalue of the variance-covariance matrix
across imputations.

The two methods will be evaluated with an estimand per scalar summary of
interest. It is assumed that when the estimates are unbiased (and
confidence valid), the algorithm is sufficiently converged. The
\(\widehat{R}\) and \(AC\) values with chain means as theta are
evaluated against the bias in univariate mean estimates (\textbf{just
for Y??}). Similarly, the performance measure for Rhat and \(AC\)
applied to chain variances is the bias in estimated standard deviations.
To evaluate the performance of Rhat and \(AC\) on the eigenvalues, we
will use bias in R squared and in the estimated regression coefficient,
the coverage rate of the CI95\% of the regression estimate, and the CI
length. \textbf{remove the term performance measure above and add that
performance measures are bias in all estimands, and cov and CIL of
regression coeff.}

@@

In general there is more bias in the estimate in conditions with a
higher proportion of missingness and or a lower number of iterations. As
required the values of o\(\widehat{R}\) R lower for conditions with a
higher number off iterations. End somewhat higher 4 conditions with a
higher percentage of missing data. However autocorrelation did not show
a decrease with increase number of iterations. After replacing the ACF
function with a manual calculation for autocorrelation the
autocorrelation values were indeed decreasing with a higher number of
iterations. Evaluation with the performance measures shows that are had
an autocorrelation are conservative: signs of non-convergence in
conditions where estimands are unbiased and or confidence valid. As
expected complete convergence (as indicated by autocorrelation equals 0;
\(\widehat{R}\) equals 1) is not observed.

@@

Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
1.01. And a t-test is performed for \(AC\) values == 0. Performance
measures are the same as above: unbiased, confidence valid estimates.

@@

Our head is equal to 1.2 is not stringent enough because the performance
measures R not unbiased and confidence fellas fell it 4 every condition
in which \(\widehat{R}\) is smaller than one point 2. At least, not when
\(\widehat{R}\) is smaller than one point two fur first occurs. The
threshold \(\widehat{R}\) is equal to one point one is somewhat
conservative with respect to the performance measures. It seems OK.
\(\widehat{R}\) is equal to 1.01 is too stringent compared to the
performance measures (\textbf{but may be necessary in more complex
complete data models???}).

@@

For empirical researchers: 1) Check trace plots for pathological non
convergence and adjust imputation model if necessary. 2) use
\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
\textbf{?} as threshold . Keep iterating until these thresholds are
reached. 3) Do not use the R function ACF. Instead compute
autocorrelations manually , see for example GitHub link. 4) Track your
own scalar summary of interest. This is advanced but explained in Van
Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this
scalar summary.

@@

If I don't include it in my study, also: Significance of \(AC\) values;
Estimate (PCA or reg. coeff.) versus Rhat and \(AC\) on the estimate

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2, ..., 100\)), and 2) varying the
missingness proportions (\(p_{miss} = .05, .25, .50, .75, .95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the initial values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

@@

\(\widehat{R}\) (\textbf{of what?? i.e., chain means and chain variances
of each variable}) is computed by implementing Vehtari et al.'s
\citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
is \(AC\) aggregated across imputations? i.e., mean in latest sim}) is
computed with function \texttt{stats::acf()}.

To estimate the quantity of scientific interest, \(Q\), we perform
multiple linear regression on each completed dataset with the function
\texttt{stats::lm()}. We obtain an estimated regression coefficient per
imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
the function \texttt{mice::pool()} to estimate u bar (\textbf{check
definition!}) according to Rubin's \citeyearpar{rubin87} rules, and
subsequently implement finite population pooling conform \citet{vink14}.

@@

Auto-correlations indicate no sign of trending within imputation chains.
In most simulation conditions \(AC\) is smaller than or about equal to
zero. Across simulation conditions, however, the autocorrelation curve
does not trend towards 0. Autocorrelation values plateau off at a value
of around .1 . This is a small positive autocorrelation, which would
indicate some trending within chains.

@@

The negative \(AC\)-values obtained in this study show no threat of
non-stationarity. However, initial dip in \(AC\)-values may have
implications for the default number of iterations in \texttt{mice}
(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
the most appropriate, since this lead to the worst convergence
(\textbf{nope, only for Rh, not \(AC\)}), as indicated by
\(\widehat{R}\) and \(AC\). Under the current specifications, \(T>20\)
would be more appropriate.

The observed dip in \(AC\) implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant.
